\label{chapter:introduction}%
%
Controlling complexity is a central goal of any programming language, especially
as software written in that language grows. The burgeoning field of
\emph{variation theory} and \emph{variational
  programming}~\cite{EW11gttse,EW11tosem,HW16fosd,CEW16ecoop,Walk14onward}
attempt to control a kind of complexity which is induced into software when many
\emph{similar yet distinct} kinds of the same software must coexist. For
example, software is ported to other platforms, creating similar, yet distinct
instances of that software which must be maintained. Such instances of variation
are ubiquitous: Web applications are tested on multiple servers; programming
languages maintain backwards compatibility and so do software libraries;
databases evolve over time, locale and data; and device drivers must work with
varying processors and architectures. Variation theory and variational
programming have been successful in small
systems~\cite{ES18diagrams,SE17fosd,MMWWK17vamos}, but have not been tested in
a performance demanding practical domain. In the words of Joe
Armstrong\cite{armstrongThesis}, ``No theory is complete without proof that the
ideas work in practice''; this is the project of this thesis, to put the ideas
of \emph{variation} and \emph{variational programming} to the test in the
practical domain of \acl{sat}.

To my knowledge there is no formal definition of a \emph{variation-aware} or
\emph{variational} system, and the terms are used interchangeably in the
research literature. For clarity, I provide the following working definitions:

\begin{definition}[Variational Problem]
  A problem is variational if separate instances of the problem share elements
  which are critical to solving the instance of the problem, and elements which
  are not shared are able to be deterministically represented.
\end{definition}
%
Consider the non-variational problem of finding the sum of a set of numbers,
such as $\kf{ns} = \set{1,2,3,5,7}$. The problem becomes variational when we
want to find the sum of $\kf{ns}$ and another sum for a related set $\prime*{ns}
= \set{1,2,3,13,17}$. \prime*{ns} shares three elements $1$, $2$, and $3$ with
$\kf{ns}$, and so finding the sum of each is a related but distinct instances of
the non-variable problem. The shared elements cannot be ignored, as this would
change the solution for each set, and the elements which are not in both sets
can be deterministically represented either by a subset or property: \setc{n}{n
  \in \kf{ns}, n \notin \prime*{ns}}.
%
Now we can define variational systems as as systems which exploits these facts
to solve variational problems.

\begin{definition}[Variational or variation-aware System]
  A system is variational or variation-aware if it uses the shared elements of a
  variational problem to solve all or some instances of the problem.
\end{definition}
%
So a variational summation of $\kf{ns}$ and \prime*{ns} should exploit the fact
that \set{1,2,3} are shared elements. A viable strategy might be to translate
the problem into two sub-problems, one to sum the shared elements and one to sum
the rest. With this procedure the summation of $\kf{ns}$ is $\sum_{n \in{} ns} n
= \sum_{s \in{} \kf{ss}} s + \sum_{n \in{} \kf{ns} \setminus{} \kf{ss}} n$ where
$\kf{ss} = \setc{s}{s \in{} \kf{ns}\ \cap{}\ \prime*{ns}}$. Since the
variational summation understands which elements are shared and which are not,
the procedure can sum the shared elements only once, and then reuse that result
for the sum of \prime*{ns}. This allows the variational summation to avoid
redundant summations over the shared terms \emph{because} we have framed the
problem as a \emph{variational} problem.

The major contribution of this thesis is the formalization of a
\emph{variational satisfiability solver}, that is, a satisfiability solver which
exploits shared terms between sets of related \acl{sat} problems to solve some
or all of those problems efficiently. In search of this goal the thesis
additionally formalizes \emph{\acl{vpl}}, and constructs a \emph{variational
  \ac{sat} solver}. In the next section I motivate the use of variation theory
and variational techniques in \acl{sat}. In addition to work on variational
\ac{sat} several other contributions are made. The thesis extends variational
\acl{sat} to variational \ac{smt}. It presents reusable techniques and
architecture for constructing \emph{variational or variation-aware} systems
using the non-variational counterparts of these systems for other domains. It
shows that, with the concept of variation, the variational \ac{smt} and \ac{sat}
solvers can be parallelized. Lastly, this thesis is the culmination of many
conversations, arguments and work from myself and many others over the last five
years. To honor their contributions I use the first-person plural ``we'', rather
than ``I'' for the remainder of the thesis.

\section{Motivation and Impact}
\input{Chapters/Introduction/Motivation}

\section{Contributions and Outline of this Thesis}
\input{Chapters/Introduction/Contributions}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../thesis"
%%% End: