Over the last two decades, satisfiability and satisfiability-modulo theory
(SAT/SMT) solvers have grown powerful enough to be general purpose reasoning
engines throughout software engineering and computer science. However, most
practical use cases of SAT/SMT solvers require not just solving a single SAT/SMT
problem, but solving sets of related SAT/SMT problems. This discrepancy was
directly addressed by the SAT/SMT community with the invention of incremental
SAT/SMT solving. However, incremental SAT/SMT solvers require end-users to hand
write a program which dictates the terms that are shared between problems and
terms which are unique. By placing the onus on end-users, incremental solvers
couple the end-users' solution to the end-users' \emph{exact} sequence of
SAT/SMT problems---making the solution overly specific---and require the
end-user to write extra infrastructure to coordinate or handle the results.

This dissertation argues that the aforementioned problems are caused from the
lack of variation as a computation concept, similar to that of a \rn{while}
loop. To demonstrate the argument, this thesis applies theory from
\emph{variational} programming to the domain of SAT/SMT solvers to create the
first variational SAT solver.
%
The thesis formalizes a variational propositional logic and specifies
variational SAT solving as a transpiler, which transpiles variational SAT
problems to non-variational SAT that are then processed by an industrial SAT
solver. It shows that the transpiler is an instance of a variational fold and
uses that fact to extend the variational SAT solver to an asynchronous
variational SMT solver. Finally, it defines a general algorithm to construct a
single variational string from a set of non-variational strings.
% I show numerous benefits gained with this approach:
% %
% A variational solver is more declarative than a SAT solver as previously
% necessary control flow operations are no longer required.
% %
% As a result, end-users need only identify the set of SAT problems of interest
% rather than identify the set \emph{and} direct the solver via the control flow
% operations.
% %
% With the concept of variation, previously difficult optimizations can be
% syntactically detected and applied before the solver runtime.
% %
% When solving a set of related problems a variational solver promotes more reuse
% of information and can thus be more performent than a non-variational solver in
% its use case.

% Furthermore, \todo{double check these}the thesis explores several extensions and
% reflect on lessons learned from implementing a large scale variation-aware
% system. It defines a general algorithm to construct a single variational string
% from a set of non-variational strings. It extends variational SAT solving to
% variational SMT solving, and with variational constructs, it shows that a
% variational SAT/SMT solver can trivially  be made asynchronous.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End: