No research is the sole product of the first author and this work is no
exception. Numerous people have contributed to it in a myriad of ways, from
supporting me as I worked, to argument and to outright collaboration.

To begin I would like to thank my partner Parisa Sadat Ataei for, well,
innumerable instances of support. Thank you for helping develop variational
propositional logic on a black board in my second year. Thank you for begin with
me during the late nights and dark times. Thank you for encouraging me to go to
Germany in my fourth year even though it hurt you to let me go. I look forward
to many more drives through the countryside with you and life outside of
graduate school.

I must also thank my advisor Eric Walkingshaw, without whom this work would not
have been possible. Thank you for being a great advisor, friend, and mentor; for
getting excited about my ideas, and for allowing me to explore but not wander. I
am honored to have had you has my advisor in graduate school. Most of all thank
you for having the soul of a researcher and for being a good person. I count
myself as lucky to have finished my PhD in a research group that was dedicated
to the craft, that treated the job as making contributions to scientific
discourse. For that I will always be grateful.

Thank you to Thomas Thüm, Paul Maximilian Bittner, and Tobias Heß. Tobias for
supporting me during my time in Germany, complete with a delivery toilet paper
during the first weeks of the pandemic and many pleasant nights arguing
politics. Thomas for too many things to recount here: hosting my stay in
Germany, complete with an apartment and aid during the beginning of a pandemic.
Putting in the time to editing drafts of our publications, making sure
through-lines were kept intact. Offering career advice, sample data,
coordinating an amazing afternoon cycling trip through the beautiful countryside
of Baden-Württemberg, and of course the morning 5km runs through Ulm; memories
which are dear to me. Paul for his many contributions to this work, including
his many comments on the work, revision of several definitions and aspects of
the logic, and his help in helping me learn how to explain the work to others.
Thank you Paul, my friend.

I would like to thank Ameya Ketkar the warlock, William Maxwell the wizard, and
Colin Shea-Blymyer the bard. Our weekly Dungeons and Dragons games are a
constant delight. Ameya and Will, thank you for helping me get the functional
programming club up and running, it is a shame that we didn't have time to
nurture it more. Ameya, thank you for your clear eyed and precise observations,
and drive to get to the top of every hill and mountain in Oregon. Will, thank
you for your rants on category theory and computational geometry. Colin, thank
you for your skill at libation, for being a charitable interlocutor, and being a
lover of wisdom. I am sad that our time in graduate school is coming to an end,
my fellow adventurers.

Last but not least I would like to thank all the regulars at Crossfit Train in
Corvallis; especially the owner Derek Eason. Thank you Derek, without you I do
not believe I would have finished this PhD. Thank you for being a constant
positive force in my life, for reminding me what normal looks like, and above
all for being an excellent coach and person. I will truly miss the people and
times at Crossfit Train; my home away from home.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End: