~\label{sec:research-plan}
\autoref{sec:prop-contr} lists the deliverables for this thesis. Each item has
been discussed in more depth in previous sections. This section summarizes these
items and provides an itemized list of each deliverable, broken down into
discrete tasks, and their corresponding time estimations. The time estimates are
guidelines with the real schedule being dictated by our publication schedule.

\subsection{Encoding strategies of \ac{vpl} formulae}
The naive and naive-with-optimizations encoding algorithms are complete. The
remaining work is a literature review for other representations of boolean
formulae which may be useful, and to implement a Huffman based encoding
algorithm and assess its performance:

\begin{enumerate}
\item Construct a list of alternative representations of boolean formulae which
  may lend itself to encoding strategies
\item\label{huffman-todo} Implement the Huffman coding algorithm over boolean
  formulae, recursively merge formulas into a \ac{vpl} formula based on a
  similarity metric.
\end{enumerate}

\textbf{Estimated time to complete: 1-2 weeks}

\subsection{A method to determine the hardness of \ac{vpl} formulae}
The time consuming parts of this item: creating a random generator suitable for
the analysis and creating a benchmark platform to test are complete. The only
items remaining are to perform the analysis.

\begin{enumerate}
\item\label{phase-todo} Review literature from random-ac{sat} community on observing phase
  transition for both random $k$-\ac{sat} and mixed-\ac{sat} problems.
\item Replicate the transition detection analysis from \autoref{phase-todo}.
\end{enumerate}

\textbf{Estimated time to complete: 1 week}


\subsection{Variational \ac{smt} Solving}
A prototype asynchronous variational \ac{smt} solver is complete and operational
but several auxiliary tasks are still required. I expect this item to require
the majority of time besides writing the the proposed thesis.

\begin{enumerate}
\item\label{ui-todo} Implement an enhanced user interface which treats the
  solver like a server rather than a batch process.
\item\label{vpl-todo} Write the formal grammar for \ac{smt}-flavored \ac{vpl}.
  The current formalization exists in Haskell code and thus must be translated
  into a publishable format.
\item\label{vmodel-todo} Write the formalization for \ac{smt}-flavored variational models. An
  identical item to \autoref{vpl-todo} only with variational models.
\item \label{solve-todo} Write the denotational semantics for \ac{smt}-flavored
  variational solving algorithm. An identical item to \autoref{vpl-todo} only
  for the solving algorithm.
\item \label{proof-todo} Construct a proof for variation preserving semantics
  using the formalizations in items 2-4.
\item\label{nano-review-todo} Review literature on nanopass compilers to
  construct a set of possible optimizations.
\item\label{opts-todo} Implement optimizations from \autoref{nano-review-todo}
\item \label{conf-todo} Implement configurations of the variational solver to
  provide an interface suitable for benchmarking.
\end{enumerate}

Estimated time to complete 1-4: 1-2 weeks

Estimated time to complete \autoref{proof-todo}: 4 weeks (concurrent with other work)

Estimated time to complete items 6-7: 2 weeks

\textbf{Total estimated time to complete: 6-8 weeks}

\subsection{Assembling Case Study Data}
Most of the work left to assemble the case study data involves refactoring and
running the forked TypeChef system on the Linux kernel; the Busybox case study
dataset is finished. Parsers for the \ac{spl} case studies and the Busybox/Linux
results are already complete and have been tested.

\begin{enumerate}
\item\label{tc-todo} Implement a database caching system in TypeChef to record \ac{sat}
  problems at scale
\item\label{run-linux-todo} Run the Linux case study to generate \ac{sat}
  problems.
\item\label{eval-linux-todo} Evaluate the Linux and Busybox case studies using
  encoding strategies from \autoref{huffman-todo}.
\end{enumerate}

Estimated time to complete \autoref{tc-todo}: 2-3 weeks (with help from
collaborator Paul Maximillian Bittner)

Estimated time to complete \autoref{run-linux-todo}: 1 week

Estimated time to complete \autoref{eval-linux-todo}: 1-2 weeks

\textbf{Total estimated time: 4-6 weeks}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:
