\label{chapter:case-studies}
%
We have formalized variational \ac{smt} and \ac{sat} solving. However, we have
yet to investigate the performance of our methods. Recall from
\autoref{chapter:introduction} that a motivating reasons for a variational
solver was that if we only compute shared terms once, then we should observe a
speedup in runtime performance when solving sets of related \ac{sat} problems
because information is reused. In this chapter, we investigate and verify these
claims.

Assessing the performance of \ac{sat} and \ac{smt} solvers is notoriously
difficult~\cite{Gent94thesat} because it depends on the input problem to the
\ac{sat} or \ac{smt} solver. The issue is related to the computational hardness
of the input. Hardness, in this domain is estimated by the ratio of clauses in
the \ac{sat} or \ac{smt} problem to the number of variables. Conceptually, if
there are many clauses and not many variables then the problem is
over-constrained and it is easy to decide \rn{unsat}, however if there are few
clauses but many variables then the problem is under-constrained and it is easy
to decide \rn{sat}. Thus, hard problems rest in a \emph{phase transition} zone
where the ratio of clauses to variables is neither over-constrained nor
\todo{cite}under-constrained.

To investigate the performance of our methods, we construct a prototype
variational solver, \vsat{} in the Haskell programming
language~\cite{Hudak:1992:RPL:130697.130699} and quantitatively compare it to
incremental and non-incremental \ac{sat} solving. Assessing the prototype in
realistic conditions is difficult as there does not exist a corpus of accepted
representative variational \ac{sat} problems. Thus, to test the prototype in as
realistic conditions as possible we utilize real-world data from a previous
study by~\citet{NMS+:GPCE18} from the \ac{spl} community.

Before we describe the datasets and resulting variational \ac{sat} problems we
first introduce some terminology from the \ac{spl} community. A software
product-line is an instance of variational software; \todo{get a source here}or
more formally it is a set of software-intensive systems that share a common,
managed set of features satisfying the specific needs of a particular market
segment or mission and that are developed from a common set of core assets in a
prescribed way.

A good example of a software product-line is the Linux kernel~\cite{linux}. The
Linux kernel is a set of core assets which devise an operating system, but the
assets are parameterized by \emph{features} which, in this case, are the Boolean
conditions of conditional-compilation statements such as \cpp{ifdef}s. To select
the particular kind of kernel to build, the Linux kernel uses the
\texttt{KConfig}~\cite{kconfig} tool to enable or disable features and thus
specify the exact kernel to build. The set of features and their dependencies
which determine the product, or in this case determine the kernel that is built
is call a \emph{feature model}\cite{KCHNP90}.

It is common to express feature models as a \ac{sat} formula where variables are
features, and dependencies are expressed using logical connectives. Thus,
reasoning about feature models with a \ac{sat} solver is an active sub-field in
\acl{spl}~\cite{BSRC10,GBT+19,BSTRC06,TAK+:SE15}. For example, a \emph{void
  analysis} uses a \ac{sat} solver to determine that a product is possible, and
a \emph{core analysis} manipulates the feature model to check that a given
feature must be \tru{} (or enabled in the \ac{spl} terminology) for every viable
product. Conceptually in this domain, if a \rn{sat} is returned from the solver
then the resulting model is an assignment of features which specifies a viable
product. If an \rn{unsat} is returned than no viable product exists given the
constraints on the feature model for the software product-line. Further analysis
on the product line are

\section{Experimental Methodology}
\input{Chapters/CaseStudies/Experimental_Methodology}
%
\section{Results and Discussion}
\input{Chapters/CaseStudies/ResultsAndDiscussion}

\section{Variational \ac{smt} Results and Discussion}
\input{Chapters/CaseStudies/Vsmt_Results}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../thesis"
%%% End: