\label{section:related-work:similar-solvers}
%
This work is most similar to the Green solver by \citet{VGD:FSE12}, which also
constructs a \ac{sat} solver that exploits shared terms and prevents redundant
computation. However, the projects differ in important ways. Visser et al.'s
solver is oriented for program analysis and does not use incremental \ac{sat}
solving. Rather, it employs heuristics to find canonical forms of sliced
programs, and caches solver results on these canonical forms in a key-value
store~\citep{redis}. In contrast, variational \ac{sat} solving is domain
agnostic, solves \ac{sat} problems expressed in \ac{vpl}, returns a variational
model, and uses incremental \ac{sat} solving.

It is also possible to view incremental \ac{sat} and \ac{smt} solvers and the
incremental \ac{sat} problem as variational systems and as a variational
problem. Both are concerned with efficiently solving instances of problems which
by definition share terms and are therefore related, and thus variational. We
provide a small history of incremental \ac{sat} here as it is related work by
being the target language of our compiler.

The incremental \ac{sat} problem was first defined by \citet{hooker_1993}, with
successive refinements of techniques by~\citet{branch-bound}, and with the
assertion stack idea developed in~\citet{kim2000solving}\todo{double check}. The
incremental \ac{sat} problem was devised as a solution to verification and
optimization problems in electronic design automation such as covering
problems~\cite{10.1145/217474.217603}, detecting delay
faults\cite{10.1145/343647.343801}, and model
checking\cite{Clarke:1986:AVF:5397.5399}. The first incremental solver to gain
traction was \texttt{SATIRE} created by \citet{10.1145/378239.379019}.

Just two years later, \citet{10.1007/978-3-540-24605-3_37} made a major advance
in incremental \ac{sat} with \texttt{MiniSat} by defining, documenting, and
popularizing the implementation techniques required for an incremental \ac{sat}
solver. \texttt{MiniSat} was the result of work on two other solver's called
\texttt{SATZOO} and \texttt{SATNIK}, and seemed to hit a sweet spot in the
design space. It simplified the existing notions of incrementality from the
state of the art incremental solvers \texttt{SATIRE} and
\texttt{PBS}~\cite{10.1145/774572.774638} and combined propagation strategies
from the \texttt{Chaff}~\cite{Moskewicz:2001:CEE:378239.379017} solver such as
conflict-driven backtracking\cite{Zhang:2001:ECD:603095.603153} and dynamic
variable ordering~\cite{Moskewicz:2001:CEE:378239.379017}. These combinations
lead to a solver that was performent, and whose implementation was small and
communicative. That same year, the first SMTLIB standard would be proposed by
\citet{SMT-LIBformat} although incremental \ac{sat} commands would not be
incorporated until the 2.0 version\cite{BarST-RR-10} in 2010.

The use of choices in the variational solvers is similar to the concept of
\textit{facets} by~\cite{austin2012multiple} and \textit{faceted execution}
by~\cite{Schmitz2018FacetedSM,Micinski2018AbstractingFE,10.1145/2465106.2465121},
in that both choices and facets syntactically demarcate terms in an object
language that must be specially handled, and yet must also operate with terms
outside of the choice or facet. Facets are very similar to choices, facets use a
label to determine branches (or alternatives in our language), facets are
synchronized by these labels, facets are treated as tree-data structures, and
facets are similarly treated as undetermined until they are reified.

\citet{10.1145/3243734.3243806} define the faceted secure execution framework
\texttt{Multef}, which tries to avoid repeated execution of non-faceted values
just as this work tries to gain performance through avoiding repeated execution
of plain values. \texttt{Multref} does this by forking executions threads when a
novel facet is encountered. This strategy avoids redundant execution before the
facet is found but still has redundant or repeated computations inside the fork.
In contrast, our methods of accumulation, evaluation, and utilization's of a
zipper succeeds in only evaluating plain terms a single time and reusing that
information across variants. The facets have been employed to policy-agnostic
programming models and information flow control~\cite{IFC}, thus our methods
might leak too much information to be useful in that domain.

However, there are other striking similarities, \citet{optimisingFacets}
improves the performance of \texttt{Multref} by defining rewrite rules which
manipulate facets similarly to the equivalence laws presented for choices in
\autoref{fig:cc:eqv}. For example, Algehed et al. removes redundant facets
through a rewrite rule called \texttt{Choice Irrelevance}, which is isomorphic
to the \rn{IDemp} rule in \autoref{fig:cc:eqv}. Another case is definition of
\emph{Squashes} which finds dead branches in nested facets. Squashes are
similarly isomorphic to our discussion of \emph{dominating choices} in
\autoref{section:vpl:semantics}.



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../thesis"
%%% End: