\label{section:related-work:variational-systems}
%
Variational \ac{sat} solving is the latest in a line of work that uses the
choice calculus to investigate variation as a computational phenomena. This body
of work ranges from data structures, to graphics, to full fledged systems such
as the system presented in this thesis. Due to the nature of variational
problems, many variational or variation-aware systems employ \ac{sat} and
\ac{smt} solvers. We collect and discuss these contributions here beginning with
variational data structures.

There is relatively little work on variational data structures.
\citet{EWC13fosd} describes variational sets and graphs. \citet{Walk14onward}
advocate for research on variational data structures and explore the trade-offs
made in ad-hoc implementations used in variational systems such as
TypeChef~\cite{KKHL:FOSD10} and SuperC~\cite{GG:PLDI12}. For this section, we
focus on recent advancements implementing performent variational stacks and
lists. The goal of variational data structures is to construct a data structure
which describes a set of non-variational data structures and supports efficient
variational operations. The variational artifact is the implementation of the
variational data structure, and the variants in this domain are the plain
versions of the data structure or plain values that result from operating on the
data structure. The challenge is to devise a variational data structure that
describes and contains the variation, and provides a set of operations to
manipulate the data structure that are as close to the performance of their
plain counterparts as possible.

A fundamental tension in this domain is exemplified by work on variational
stacks by~\citet{MMWWK17vamos}. Meng et al.\ defined two kinds of variational
stacks: a stack of choices, or a choice of stacks. Their analysis on a general
implementation strategy was inconclusive. They found that depending on the
implementation strategy runtime performance could be affected by as much as
20\%. Furthermore, the variation in their experiment is coarse grained, \ie{},
the sharing ratio is high. Thus, Meng et al.\ used heuristics (optimizations in
their paper) which further improved performance for both kinds of variational
stacks by 43\%. Using heuristics was also found to be a successful strategy in
Meinicke's PhD dissertation which we address below.

The work on variational stacks yields an alternative implementation strategy for
variational \ac{sat} solvers. We have carefully designed our variational
\ac{sat} and \ac{smt} solvers to use a plain base solver. We could have done
otherwise and implemented a variational solver directly. With variational stacks
the variational solver could use a variational assertion stack and we would
avoid the need for a zipper in choice removal. Such an implementation is worth
considering although by developing an independent solver we lose any benefits
brought by the \ac{sat}/\ac{smt} communities and lose the general recipe for
constructing a variation-aware system \emph{using} its plain counterpart.

Similar to variational stacks, \citet{SE17fosd} successfully implemented
variational lists. Smeltzer and Erwig devise six implementations of variational
lists with one implementation, the \emph{suffix list} coming from previous
work~\cite{EW11gttse}. Smeltzer and Erwig's study leads to some surprising
results. Out of their six implementations they found that for some
implementations, simple functions such as \texttt{head} (which returns the first
element of the list) are slower than the brute force counterpart because the
implementation may be required to traverse the whole list to resolve the
variation. However, they do conclude that one implementation, the \emph{segment
  list}, yields reasonable performance given the data in their study. The
segment list is an interesting result as the idea behind the design is to encode
variation as a \emph{sequence of segments}, where a segment is either a choice
or a sequence of plain elements. This idea should sound familiar, as
accumulation and symbolic values are essentially pointers to sequences of plain
terms. Smeltzer and Erwig also observe that the sharing ratio has a measurable
impact on performance (a finding we also observed) and thus minimizing or
manipulating choices to increase the ratio is important, a result that has also
been observed in \ac{spl} by~\citet{ARW+:ICSE13} and~\citet{KRE+:FOSD12}.

In addition to data structures there has been research on applications of the
choice calculus to graphics~\citep{ES18diagrams}, type
systems~\citep{CCEW18popl,CCW18icfp,CEW:TOPLAS14,CEW12icfp}, and error
messages~\citep{CES17jvlc,CE14popl,CEW12icfp,CES14hcc}. For the remainder of
this section we focus on variational or variation-aware systems.

This work is not the first to construct a variational or variation-aware system.
Notably, \citet{LKA+:ESECFSE13} produced TypeChef, which used the choice
calculus and variational data structures to type check every possible Linux
kernel. Constructing a variational parsing~\cite{KGR+:OOPSLA11}, a variational
lexer~\cite{Kastner11partialpreprocessing}, type
system~\cite{LKA+:ESECFSE13,KOE:OOPSLA12} and control-flow and data-flow
analyses~\cite{LKA+:ESECFSE13}. Similarly, \citet{GG:PLDI12} variationally parse
the Linux kernel by using variational data structures and \emph{choice nodes} in
the abstract syntax tree. TypeChef is notable for several reasons: its
implementation is a direct inspiration for our baseline algorithm \vTop{} which
uses an incremental \ac{sat} solver but only exhibits sharing \emph{before} a
choice is discovered. This kind of sharing, called \emph{prefix sharing} by
\citet{SE17fosd} is the de-facto standard in software product-line applications
which employ incremental \ac{sat} solvers. Given the results of this thesis,
large performance gains are possible if our results are representative with the
use of a variational \ac{sat} or \ac{smt} solver. TypeChef is also notable for
its two step approach, first it parses source code to find \cpp{ifdef}
annotations, and stores these in files called \emph{presence conditions}.
Presence conditions are isomorphic to variation contexts, both are \pl{}
formula's over dimensions (or conditions of the \cpp{ifdef}) which determine a
variant. Using the presence conditions, TypeChef annotates choice nodes to
determine which variant the leaves of the choice node belong. Then TypeChef
extends the symbol table of a C program to contain types which are conditional
based on the presence conditions. This allows a variable's type to change from
one variant to another. Each type checking operation is then lifted to handle
the variational cases and then type checking checks the variation-aware types to
ensure every each variant type checks. Similar to our use of variation contexts,
TypeChef allows a \emph{variability model} which specifies variants that should
be type checked by conjoining the model with the presence conditions.

In his PhD dissertation, \citet{M14,JensDebugging} constructs a variational
interpreter called VarexJ and a variational bytecode transformer called VarexC,
to achieve a variational execution and debugging framework. The framework tries
to maximize sharing in two ways: First, it directly utilizes the choice calculus
to represent local points of variation and achieves a \emph{fine-grained
  approach}, this allows the framework to share program states and keep a
unified heap. Second, the framework achieves instruction-level sharing among
control-flows between variants. It achieves this by implementing a variational
scheduler, which seeks to order the execution of program statements to optimize
sharing. We achieve this same effect through the interaction between
accumulation, evaluation and choice removal with the wrapped primitive
operations. Interestingly, Meinicke identifies redundant \ac{sat} calls as a
major bottleneck in the variational execution framework. Specifically they
determine redundant \rn{check-sat} calls as the most expensive operation in
their system. To reduce the redundant calls, the variational execution framework
caches calls to the solver, thus only employing the solver for new queries. This
technique proved effective for their domain and effectively eliminated the
bottleneck. Lastly, in his PhD dissertation Chu-Pan Wong used VarexJ to do
speculative mutation testing and automated program repair~\cite{ChuPanThesis}.
While not a variational system, Wong's work is notable for using a variational
system, defining variational expectation traces, and employing a \ac{sat} solver
to find interesting mutants to test.

% Lastly, Meinicke's work was put to use by \citet{ChuPanThesis} to induce
% \emph{speculative variation} for automatic program repair and mutation testing.
% We only touch on this work briefly as it applies a variational system rather
% than creating one. However, the work utilizes the aforementioned variational
% execution framework and a \ac{sat} solver to reduce the search space of
% interesting mutants to test.

Lastly, choice calculus has been successfully applied to databases to construct
a complete approach for variational databases including a variational database
management system, a variational query language, and variational tables.
\citet{ATW17dbpl,ATW18poly} add choices to relational algebra to define a
variational query language for a variational relational database. The
variational query language serves as the variational artifact similar to the
role of \ac{vpl} in the variational \ac{sat} and \ac{smt} solvers. Ataei et al.
specifically choose to avoid adding choices to variational tables, instead
opting to apply \emph{annotations} to the table schema, table attributes, and
table tuples. The annotations are \pl{} formulas and are derived from the
dimensions of choices in the variational query language. Annotations that are
attached to an aspect of the database, such as a schema, attribute, or tuple are
called \emph{presence conditions} following the work on TypeChef. Annotations
which are not attached but describe possible variants are called \emph{feature
  expressions}.

This careful design has several desirable properties: The separation between the
variational aspects of the system and the database engine allows the database
engine to remain mostly unchanged. Thus, Ataei et al. avoid implementing low
level details such as a variational \texttt{B+-Tree} or file system. However the
system is still memory efficient: Elements which are shared between variants are
represented a single time in the database. To realize an element is shared, a
\ac{sat} solver is called on the presence conditions for that element. For
example, imagine an attribute that belongs to two variants $\kf{A}$ and
$\kf{B}$, to encode that this attribute belongs to these two variants is
expressed in the presence condition as a disjunction, $\kf{A \vee B}$. Thus,
Ataei et al.'s system is a mixed approach; the query language embeds choices to
explicitly represent local points of variation. The underlying object language
(the database in this case) lacks a primitive operation to handle variation such
as the \rn{push}/\rn{pop} commands in our work. Hence, Ataei et al. choose to
\emph{realize} variation in the database through indirection based on
annotations and \ac{sat} solving. Thereby enabling a full fledged variational
database without requiring substantive changes to the entire database
implementation. Therefore, Ataei et al.'s system is more expressive than the
variational \ac{sat} and \ac{smt} solver's presented in this thesis because it
can express dependencies between variants through presence conditions, while our
approach is limited to express dependencies by nesting choices.

Notably, a major limitation of our method for variational \ac{sat} and \ac{smt}
solving is that it \emph{requires} that all points of variation to be known
\emph{before} running the solver. This is a direct consequence of \ac{vpl}; by
construction one can only make a \ac{vpl} formula if a point of variation is
known. If one does not know, or needs to discover the points of variation at
runtime then the \ac{vpl} formula cannot be constructed. This limitation is a
significant difference from incremental solvers. We return to this point in
\autoref{section:conclusion:future-work}, but using variational \ac{sat} solvers
effectively in these domains is an open research question.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../thesis"
%%% End: