\label{chapter:vsat}

This chapter presents the variational satisfiability solving algorithm as a
compiler. First we formalize the variational satisfiability problem. Next we
present a general overview of the solving algorithm in
\autoref{section:vsat:approach}. \autoref{section:vsat:approach} describes the
algorithm in terms of communication between four subsystems, introduces the
concept of \emph{variational cores}, and the intermediate language that each
subsystem operates on. Each of these subsystems process the intermediate
representation in a combination of phases which are covered in detail in the
following sections, including examples of the compiled \acl{smtlib} output. We
conclude the chapter with \autoref{section:vsat:formalization}.
\autoref{section:vsat:formalization} presents inference rules to specify the
behavior of an abstract variational satisfiability solver by formalizing the
interplay between these phases, and thus the interplay between the subsystems.

Now that we have defined \ac{vpl}, we can define the variational satisfiability
problem. Like the Boolean satisfiability problem, the variational satisfiability
problem is concerned with checking satisfiability, only in this case we want to
check satisfiability of all plain variants. Thus, we define the problem as
finding a partition in the configuration space of a formula in \ac{vpl}.

\begin{definition}[Variational Satisfiability Problem]
  For a \ac{vpl} formula \fV{}. Let $\configuration{}^{*}$ be the set of all
  total configurations of \fV{}. Find a partition $\kf{(S,U)}$ of
  $\configuration{}^{*}$ such that \sem[\configuration{}]{\fV{}} is satisfiable
   $\Leftrightarrow \configuration{} \in \kf{S}$.
\end{definition}

We say a \ac{vpl} formula \fV{} \emph{is satisfiable} if $\kf{S} \neq
\varnothing$, \fV{} is \emph{totally satisfiable} if $\kf{U} = \varnothing$, and
\fV{} is \emph{unsatisfiable} if $\kf{S} = \varnothing$. Rather than return the
partition of all total configurations, our variational satisfiability algorithm
defines and returns variational models. We define these models in
\autoref{section:vsat:models}.

\section{General Approach}
\input{Chapters/Vsat/Approach}

\section{Derivation of a Variational Core}
\input{Chapters/Vsat/VarCore}

\section{Solving the Variational Core}
\input{Chapters/Vsat/Solving}

\section{Variational Models}
\input{Chapters/Vsat/Models}

\section{Formalization}
\input{Chapters/Vsat/Formalization}

% \section{Variation Preservation}
% \input{Chapters/Vsat/Variation_Preservation}
%
% Cutting the proof of variation preservation, it needs another pass between
% Eric and I and there just isn't time

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../thesis"
%%% End:
